Executive Constraints and Civil Conflict Onset

### 2020-02-24 paper.Rmd
Add GP explainer + further edits to research design

Since the end of World War II, civil conflict has been the deadliest
and most prevalent form of conflict. In that time period, upwards to
two-thirds of all countries have experienced at least one minor
insurgency and approximately one-third of all countries have
experienced a major civil war with at least 1,000 battle deaths
[@Pettersson2019]. Given the impact of civil conflict and the
increasing availability of conflict related data, the last several
decades has witnessed a veritable explosion in the number of
quantitative studies investigating the causes of civil conflict
[@Blattman2017]. Scholars have proposed numerous theories ranging from
motivation-based arguments --- such as, for example, the role of
horizontal inequalities [@Cederman2011] --- to opportunity-based
explanations centered on economic factors [@Collier2012; @Collier2009]
as well as the political and structural conditions underlying state
weakness [@Hegre2001; @Fearon2003]. This latter strain of research in
particular has spurred extensive work into examining the
distinguishing features of weak states and whether political
institutions play a significant role in the risk of civil conflict.
Several studies in this vein have argued in favor of institutions by
utilizing disaggregated measures to also investigate the specific
types of institutions relevant to the study of civil conflict
[@Taydas2010; @Hegre2015a; @Bartusevic2018].

@@

To provide a more nuanced picture, an ongoing strain of research has
been dedicated to disaggregating measures of democracy into either
finer regime type categories or specific measures of individual
political institutions. The theories offered tend to follow a general
formula whereby formal lack or poor quality of some institution(s),
measured either directly or proxied by regime type, undermines the
capacity of the state to prevent armed insurgency and/or incentivises
opposition violence by restricting the opportunities for political
participation and increasing political discontent. For example,
@Carey2007 examines the executive selection process of Sub-Saharan
Africa and argues that executive elections, even when they are of poor
quality, reduce the risk of large-scale violent dissent. @Fjelde2010a
breaks down authoritarian regime types in Africa attributing the
significant variation in conflict risk to the differing institutional
configurations affecting a regime's ability to coerce or co-opt the
opposition. The conclusion is that previous indeterminate results
regarding the link between political institutions and conflict risk
may have been hamstrung by aggregate measures which poorly reflect the
within-group variation of political regimes. In a larger sample
including democratic polities, @Goldstone2010 create a regime
classification scheme based on measures for openness of executive
recruitment and competitiveness of political participation, as well as
factionalism in order to distinguish between partial democracies. They
offer a theory of civil conflict occurring in countries where elite
divisions undermine the capacity of the state to defeat an insurgency
and claim that a relatively parsimonious model including their regime
classification offers better predictive accuracy than previous
analyses and provides evidence that political institutions are a
dominant driving factor for both political instability and civil
conflict. Finally, @Bartusevic2018 construct a regime typology focused
specifically on electoral features, which are argued form the
essential core of the concept of democracy. Where there are
constraints on electoral contestation, the authors contend that such
constraints restrict the opportunities for non-violent political
participation thereby legitimising anti-government violence and
self-selecting recruits into rebel groups willing to engage in armed
conflict. Evidence is presented which the authors claim supports not
just the significance of regime type, but a negative monotonic
relationship between democracy and the risk of civil conflict.

@@

[^model]: The model code is written in Stan [@Carpenter2017] and was
    run with the `CmdStan` interface [@Team2019a] using four parallel
    chains at 2000 iterations each with a warmup of 1000. Convergence
    was assessed according to the potential scale reduction factor
    $\hat{R}$ and effective sample size (ESS) [@Vehtari2019] as well
    as standard Stan diagnostics. The full code is available in
    [Appendix A](#appendix-a) and a fully reproducible pipeline
    compromised of replication scripts is available at
    [https://github.com/jsks/ma-thesis](https://github.com/jsks/ma-thesis). A
    simulation analysis assessing the model and it's ability to
    recover simulated true values is available in [Appendix B](#appendix-b).

@@

I model civil conflict onset as a multilevel regression where onset is
a function of a set of predictors, including executive constraints and
an array of control variables, together with several grouping factors
included as partially pooled varying intercepts. The units of analysis
are all independent country-years from 1951 until 2017, as determined
by data availability, with *I* countries indexed by $i = 1 \dotsc I$
and *T* years indexed by $t = 1 \dotsc T$. The dependent variable,
conflict onset, is a Bernoulli distributed binary variable where 1
denotes the onset of at least one new civil conflict in a given
country-year. The chosen link function transforming the linear
regression output into the probability of conflict onset, $\nu_{it}$,
is the inverse logistic such that $\nu_{it} \in [0, 1]$. [^invlogit]

[^invlogit]: Technically, the inverse logistic transforms values to
    the open interval, $(0, 1)$, but due to overflow/underflow in
    floating point calculations by computers the endpoints can be
    observed.

@@

where **X** is a vector of predictors, $\beta$ the vector of
regression slopes, and $\zeta^k$ the set of varying intercepts for *K*
grouping factors indexed by $j_k$. For the present analysis there are
only two grouping factors, year and country, to account for
cross-sectional unmodeled heterogeneity between countries and over
time [@Shor2007]. All predictors are also lagged by one year to
control for time endogeneity and standardized by substracting the mean
and dividing by the standard deviation for each variable to aid in
computation as well as comparison of the final $\beta$ estimates.

I place weakly informative priors on all parameters to allow the
posterior estimates to be largely determined by the observed data
while avoiding the more pathological values and hard constraints that
can arise from uninformative priors. For the slopes and population
intercept, given that the input data has been standardized, this
translates to:
\begin{align*}
    \alpha & \sim \mathsf{Normal}(0, 5) \\
    \beta & \sim \mathsf{Normal}(0, 2.5)
\end{align*}

For the varying intercepts modeling group-specific variability, I use
hierarchical priors to provide a regularizing effect against
overfitting [@Gelman2007].
\begin{align*}
    \zeta^k_j & \sim \mathsf{Normal}(0, \sigma^k) \\
    \sigma^k & \sim \mathsf{HalfCauchy}(0, 1)
\end{align*}

@@

Finally, in the case of the legislative variables there is the added
issue of missing observations for country-years lacking a
legislature.[^missing] For those observations, given the centrality of
the legislature as a potential constraint against the executive, the
latent factor should ideally be lower conditional on the other
manifest variables. However, since the manifest variables are
themselves unbounded latent variables, minimal values cannot be
imputed without a nonlinear transformation to a bounded scale. I thus
take a cue from *v2x_horacc* and include into the model a dichotomized
version of *v2lgbicam*, how many chambers does the legislature
contain, denoted by the Bernoulli distributed variable, $w$, where
country-years with a sitting legislature are given a $1$ and
country-years where the legislature is missing or has been suspended,
$0$. The probability of a legislature being present, $\pi_i$, is then
a linear function of the latent factor, $\theta$, transformed by the
inverse logistic such that $\pi_i \in [0, 1]$. The practical effect is
that where there is data for the legislative manifest variables, the
latent factor is informed by the measured institutional quality of the
legislature. Where data is missing due to the suspension or lack of a
legislature, the latent factor is down weighted by the linear function
predicting $w$, conditional on the observed values of the remaining
manifest variables. The resulting model specification, including
priors, is as follows:
\begin{align*}
    w_i & \sim \mathsf{Bernoulli}(\pi_i) \\
    \pi_i & = \mathsf{logit}^{-1}(\eta + \delta \theta_i) \\
    \eta & \sim \mathsf{Normal}(0, 5) \\
    \delta & \sim \mathsf{LogNormal}(0, 1)
\end{align*}

@@

[^wbgi]: Furthermore, a common issue with many measures of
    bureaucratic quality is missingness.  For example, the World Bank
    governance indicator on government effectiveness extends back only
    until 1996 (TODO: cite). Given that measures of institutional
    quality often change fairly slowly, punctuated only by abrupt
    events such as coups or revolutions, an extended time series is
    desirable. Furthermore, in the specific case of bureaucratic
    quality, there is the added issue of endogeneity. An unconstrained
    executive may seek to promote government corruption and undermine
    both meritocratic recruitment into the bureaucracy as well as its
    impartiality.

To capture the relative ease of mobilizing fighters, especially in
terms of being able to offer economic selective benefits, I include
GDP per capita. Specifically, I use the natural log of the $RGDP^o$
variable, constant price real GDP per capita, from Penn World Tables
(PWT) v9.1 (1950-2017) to allow for both cross-country and
cross-temporal comparisons of GDP per capita [@Feenstra2015]. The
justification for taking the natural log is that the marginal effect
of the variable should decrease with higher values. In other words,
the difference between the effects of relatively high levels of GDP
per capita should be significantly smaller than the difference between
the effects of low levels of GDP per capita. Finally, I also include
an interaction term between GDP per capita and the latent measure for
executive constraints to capture the potential conditional
relationship.

In addition to the previous indicators, I also include a number of
standard covariates drawn largely from previous literature. The list
of indicators are not meant to provide an all encompassing theory of
civil conflict, but rather roughly capture the opportunity structures
for rebellion.  This includes a single year lagged measure of GDP per
capita growth from PWT in order to capture short term economic shocks,
which may affect both the strength and legitimacy of the state
[@Bazzi2014]. I also include a measure of mean elevation from the
GROWup dataset [@Girardin2015] to capture terrain advantageous for
conducting an insurgency [@Fearon2003] as well as a measure of
population density derived from population data from PWT and the area
in square kilometers for each country from GROWup. Countries with a
higher population density may require a more extensive and competent
central administration and police force in order to suppress violent
dissent [@Fearon2003]. There's also the possibility that a high
population density increases the competition for resources. If the
state is identified as the source of a particular scarcity, the risk
of civil conflict may increase. Similar to GDP per capita, I also
transform mean elevation and population by the natural log with the
expectation that the scale increase should diminish as values get
larger. Finally, I include measures of ethnic fractionalization in the
form of the number of political relevant ethnic groups and the percent
of the population facing discrimination. Where there are significant
ethnic cleavages and state sponsored discrimination, ethnic ties can
facilitate collective action thereby lowering the mobilization costs
of rebel groups [@Cederman2011].

To control for the time endogeneity of conflict --- past and current
civil conflicts can impact the present risk of conflict onset as well
as the current level of executive constraints --- I include a dummy
variable for ongoing conflict [^time] and the natural log of peace
years [^peace_yrs] [@Hegre2015a]. To account for spillover effects, I
include a dummy variable for whether a neighbouring country is
currently experiencing a civil conflict. One notable lack is a general
measure of political instability, which I omit since it could be
highly correlated with executive constraints and would potentially
bias results, at least for years of crisis where the risk of civil
conflict is highest. I do, however, include a dummy variable for the
independence year since newly sovereign states may be particularly
weak.

[^time]: To be more specific, civil conflicts can be highly
    destabilizing and severely affect the governing and military
    capacity of a state, increasing the risk for an additional
    conflict onset. Moreover, in times of particularly intense
    conflict, wartime powers may be granted to the executive bypassing
    normal institutional constraints. Thus, ongoing conflict is a
    potential confounder since it can influence both the probability
    of the outcome as well as the executive constraints
    measure.

[^peace_yrs]: The peace years variable is calculated as the number of
    years since the last conflict-year, including conflict-years from
    extrasystemic conflicts. For countries left censored due to the
    dataset starting date, I set the start of the peace year
    calculation to 1946, the end of World War II. The rationale is
    that World War II was a significant international event that
    reshaped the world order and even had domestic consequences for
    countries not directly involved. For newly independent countries,
    I use the independence date as the start year.

@@

The $\delta$ and $\eta$ parameters correspond to the slope and
intercept in the linear predictor for the presence of a legislature,
*v2lgbicam*. The effect of their inclusion can be seen in figure
\ref{fig:bicam}. The latent factor, $\theta$, follows the
directionality of the manifest variables, greater values indicate a
more constrained executive and lower values vice versa. The density
plots indicate that on average, for country-years where there is no
legislature, *i.e.* $v2lgbicam$ is zero, the latent factor is lower,
although still conditional on the values for the remaining manifest
variables. Where the legislature is present, $\theta$ is distributed
according to the $Normal(0, 1)$ prior.

>>>>

Since the end of World War II, civil conflict has been the deadliest
and most prevalent form of conflict. In that time period, upwards to
two-thirds of all countries have experienced at least one minor
insurgency and approximately one-third of all countries have
experienced a major civil war with at least 1,000 battle deaths
[@Pettersson2019]. Given the impact of civil conflict and the
increasing availability of conflict related data, the last several
decades has witnessed a veritable explosion in the number of
quantitative studies investigating the causes of civil conflict
[@Blattman2017]. Scholars have proposed numerous theories ranging from
motivation-based arguments --- such as, for example, the role of
horizontal inequalities [@Cederman2011] --- to opportunity-based
explanations centered on economic factors [@Collier2004; @Collier2009]
as well as the political and structural conditions underlying state
weakness [@Hegre2001; @Fearon2003]. This latter strain of research in
particular has spurred extensive work into examining the
distinguishing features of weak states and whether political
institutions play a significant role in the risk of civil conflict.
Several studies in this vein have argued in favor of institutions by
utilizing disaggregated measures to also investigate the specific
types of institutions relevant to the study of civil conflict
[@Taydas2010; @Hegre2015a; @Bartusevic2018].

@@

To provide a more nuanced picture, an ongoing strain of research has
been dedicated to disaggregating measures of democracy into either
finer regime type categories or specific measures of individual
political institutions. The theories offered tend to follow a general
formula whereby formal lack or poor quality of some institution(s),
measured either directly or proxied by regime type, undermines the
capacity of the state to prevent armed insurgency and/or incentivises
opposition violence by restricting the opportunities for political
participation and increasing political discontent. For example,
@Carey2007 examines the executive selection process of Sub-Saharan
Africa and argues that executive elections, even when they are of poor
quality, reduce the risk of large-scale violent dissent. @Fjelde2010
breaks down authoritarian regime types in Africa attributing the
significant variation in conflict risk to the differing institutional
configurations affecting a regime's ability to coerce or co-opt the
opposition. The conclusion is that previous indeterminate results
regarding the link between political institutions and conflict risk
may have been hamstrung by aggregate measures which poorly reflect the
within-group variation of political regimes. In a larger sample
including democratic polities, @Goldstone2010 create a regime
classification scheme based on measures for openness of executive
recruitment and competitiveness of political participation, as well as
factionalism in order to distinguish between partial democracies. They
offer a theory of civil conflict occurring in countries where elite
divisions undermine the capacity of the state to defeat an insurgency
and claim that a relatively parsimonious model including their regime
classification offers better predictive accuracy than previous
analyses and provides evidence that political institutions are a
dominant driving factor for both political instability and civil
conflict. Finally, @Bartusevic2018 construct a regime typology focused
specifically on electoral features, which are argued form the
essential core of the concept of democracy. Where there are
constraints on electoral contestation, the authors contend that such
constraints restrict the opportunities for non-violent political
participation thereby legitimising anti-government violence and
self-selecting recruits into rebel groups willing to engage in armed
conflict. Evidence is presented which the authors claim supports not
just the significance of regime type, but a negative monotonic
relationship between democracy and the risk of civil conflict.

@@

[^model]: The model code is written in Stan [@Carpenter2017] and was
    run using the `CmdStan` interface version 2.22.1 [@Stan2020] with
    four parallel chains at 2000 iterations each with a warmup
    of 1000. Convergence was assessed according to the potential scale
    reduction factor $\widehat{R}$ and effective sample size (ESS)
    [@Vehtari2019] in addition to standard Stan diagnostics. All
    pre/post-processing was done using R 3.6.2 [@Core2019]. The Stan
    model code is available in [Appendix A](#appendix-a) and the full
    project in the form of a reproducible pipeline, including the
    Rmarkdown source for this manuscript, is available at
    [https://github.com/jsks/ma-thesis](https://github.com/jsks/ma-thesis).
    Finally, a simulation analysis assessing the model and it's
    ability to recover simulated true values is available in [Appendix
    B](#appendix-b).

@@

I model civil conflict onset as a multilevel regression where onset is
a function of a set of predictors, including executive constraints and
an array of control variables, together with several random
effects. The units of analysis are all independent country-years from
1951 until 2017, as determined by data availability, with *I*
countries indexed by $i = 1 \dotsc I$ and *T* years indexed by $t = 1
\dotsc T$. The dependent variable, conflict onset, is a Bernoulli
distributed binary variable where 1 denotes the onset of at least one
new civil conflict in a given country-year. The chosen link function
transforming the linear regression output into the probability of
conflict onset, $\nu_{it}$, is the inverse logit such that $\nu_{it}
\in [0, 1]$. [^invlogit]

[^invlogit]: Technically, the inverse logit transforms values to
    the open interval, $(0, 1)$, but due to overflow/underflow in
    floating point calculations by computers the endpoints can be
    observed.

@@

where **X** is the predictors, $\beta$ the vector of regression
slopes, and $\zeta^k$ the *K* sets of varying intercepts indexed by
$j_k$. All predictors are lagged by one year to control for time
endogeneity and standardized by subtracting the mean and dividing by
the standard deviation for each variable to aid in computation as well
as comparison of the posterior $\beta$ estimates.

For the present analysis there are three sets of varying intercepts,
or random effects, included in the model. This includes two partially
pooled grouping factors, year and country, to account for
cross-sectional unmodeled heterogeneity between countries and over
time [@Shor2007], and the number of peace years since the last
civil conflict incidence per country to control for the lingering
endogenous effects of violence on the likelihood that additional
conflict breaks
out.

I place weakly informative priors on the population intercept and
slope parameters to allow the posterior estimates to be largely
determined by the observed data while avoiding the more pathological
values and hard constraints that can arise from uninformative
priors. Given that the input data has been standardized, this
translates to:
\begin{align*}
    \alpha & \sim \mathsf{Normal}(0, 5) \\
    \beta & \sim \mathsf{Normal}(0, 2.5)
\end{align*}

For the varying intercepts modeling year and country specific
variability, I use hierarchical priors to provide a regularizing
effect against overfitting [@Gelman2007].
\begin{align*}
    \zeta^{year}_t & \sim \mathsf{Normal}(0, \sigma^{year}) \\
    \zeta^{country}_i & \sim \mathsf{Normal}(0, \sigma^{country}) \\
    \sigma^{year}, \sigma^{country} & \sim \mathsf{Half-Cauchy}(0, 1)
\end{align*}

Finally, for peace years I take a non-parametric approach and use
gaussian process priors with an exponentiated quadratic kernel. This
treats the number of peace years as an infinitely continuous dimension
along which the covariance between any two given points is a function
of their euclidean distance. Where that distance is smaller there is a
higher correlation. Compared to other approaches to modeling time
dependence this avoids the parametric assumption of a proportional
rate in exponential decay functions, the difficulty in selecting
interaction terms to account for non-proportionality for cubic
polynomial approximation, and the somewhat arbitrary selection of
knots for splines [*c.f.* @Raknerud1997; @Carter2010;
@Beck1998]. Given the finite set of observed peace years in the
dataset, the specification is as follows:
\begin{align*}
    \zeta^{peace\_years} & \sim \mathsf{MultiNormal}(0, K) \\
    K_{ij} & = \eta^2 \mathsf{exp}(\minus \frac{1}{2 \rho^2} D_{ij}^2)
\end{align*}

where for $P$ unique peace years, $K \in \mathbb{R}^{P \times P}$ is
the covariance matrix, $D \in \mathbb{R}^{P \times P}$ the distance
matrix, and $i$ and $j$ the column and row indexes.[^positive] The
hyperparameter, $\rho$, is the *length-scale* and corresponds roughly
to the number of years before two peace years become uncorrelated. The
hyperparameter, $\eta$, is the *marginal standard deviation* which
determines the overall variation of the output intercepts
[@Rasmussen2005].

[^positive]: To ensure that the covariance matrix, $K$, is positive
    definite a small positive term is added to the diagonal.

I select a weakly informative prior for $\eta$ that is skeptical of
large values given that $\nu$ is on the log-odds scale. For $\rho$ I
use an informative prior that constrains an overwhelming majority of
the prior probability mass to a *length-scale* between 2 and 12, which
provides a strong smoothing effect while aiding in computation by
avoiding unreasonable values [@Betancourt2017].
\begin{align*}
    \eta & \sim \mathsf{Half-Normal}(0, 1) \\
    \rho & \sim \mathsf{InvGamma}(7.30124, 30.0086)
\end{align*}

@@

Finally, in the case of the legislative variables there is the added
issue of missing observations for country-years lacking a
legislature.[^missing] For those observations, given the centrality of
the legislature as a potential constraint against the executive, the
latent factor should ideally be lower conditional on the other
manifest variables. However, since the manifest variables are
themselves unbounded latent variables, minimal values cannot be
imputed without a nonlinear transformation to a bounded scale. I thus
take a cue from *v2x_horacc* and include into the model a dichotomized
version of *v2lgbicam*, how many chambers does the legislature
contain, denoted by the Bernoulli distributed variable, $w$, where
country-years with a sitting legislature are given a $1$ and
country-years where the legislature is missing or has been suspended,
$0$. The probability of a legislature being present, $\pi_i$, is then
a linear function of the latent factor, $\theta$, transformed by the
inverse logit such that $\pi_i \in [0, 1]$. The practical effect is
that where there is data for the legislative manifest variables, the
latent factor is informed by the measured institutional quality of the
legislature. Where data is missing due to the suspension or lack of a
legislature, the latent factor is down weighted by the linear function
predicting $w$, conditional on the observed values of the remaining
manifest variables. The resulting model specification, including
priors, is as follows:
\begin{align*}
    w_i & \sim \mathsf{Bernoulli}(\pi_i) \\
    \pi_i & = \mathsf{logit}^{-1}(\eta + \delta \theta_i) \\
    \eta & \sim \mathsf{Normal}(0, 5) \\
    \delta & \sim \mathsf{LogNormal}(0, 1)
\end{align*}

@@

[^wbgi]: Furthermore, a common issue with many measures of
    bureaucratic quality is missingness.  For example, the World Bank
    governance indicator on government effectiveness extends back only
    until 1996 [@Kaufmann2010]. Given that measures of institutional
    quality often change fairly slowly, punctuated only by abrupt
    events such as coups or revolutions, an extended time series is
    desirable. Furthermore, in the specific case of bureaucratic
    quality, there is the added issue of endogeneity. An unconstrained
    executive may seek to promote government corruption and undermine
    both meritocratic recruitment into the bureaucracy as well as its
    impartiality.

To capture the relative ease of mobilizing fighters, especially in
terms of being able to offer economic selective benefits, I include
GDP per capita. Specifically, I use the natural log of the $RGDP^e$
variable, constant price real GDP, from Penn World Tables (PWT) v9.1
(1950-2017) to allow for both cross-country and cross-temporal
comparisons of GDP per capita [@Feenstra2015].[^gdp] The justification
for taking the natural log is that the marginal effect of the variable
should decrease with higher values. In other words, the difference
between the effects of relatively high levels of GDP per capita should
be significantly smaller than the difference between the effects of
low levels of GDP per capita. Finally, I also include an interaction
term between GDP per capita and the latent measure for executive
constraints to capture the potential conditional relationship.

[^gdp]: Prior to taking the natural log, the GDP measure is of course
    transformed to GDP per capita by dividing by the total population,
    which is also available in PWT.  Note, PWT has two real GDP
    measures available: $RGDP^o$, or output side real GDP, and
    $RGDP^e$, or expenditure side real GDP. The former roughly
    captures a country's productive capacity, while the latter is
    better suited to measuring living standards [@Feenstra2015]. I use
    $RGDP^e$ since I would argue that it serves as a better proxy for
    the potential labour costs for rebel leaders.

In addition to the previous indicators, I also include a number of
standard covariates drawn largely from previous literature. The list
of indicators are not meant to provide an all encompassing theory of
civil conflict, but rather roughly capture the opportunity structures
for rebellion.  This includes a single year lagged measure of GDP per
capita growth from PWT in order to capture short term economic shocks,
which may affect both the strength and legitimacy of the state
[@Bazzi2014]. I also include a measure of mean elevation from the
CShapes dataset [@Weidmann2010] as made available through the GROWup
data platform [@Girardin2015] to capture terrain advantageous for
conducting an insurgency [@Fearon2003]. In addition, I employ a
measure of population density derived from population data from PWT
and the area in square kilometers for each country from
CShapes. Countries with a higher population density may require a more
extensive and competent central administration and police force in
order to suppress violent dissent [@Fearon2003]. There's also the
possibility that a high population density increases the competition
for resources. If the state is identified as the source of a
particular scarcity, the risk of civil conflict may increase. Similar
to GDP per capita, I also transform mean elevation and population
density by the natural log with the expectation that the scale
increase should diminish as values get larger. Finally, I include
measures of ethnic fractionalization in the form of the number of
political relevant ethnic groups and the percent of the population
facing discrimination as taken from the Ethnic Power Relations dataset
[@Vogt2015]. Where there are significant ethnic cleavages and state
sponsored discrimination, ethnic ties can facilitate collective action
thereby lowering the mobilization costs of rebel groups
[@Cederman2011].

As mentioned previously, to control for time dependence --- past and
current civil conflicts can impact the present risk of conflict onset
as well as the current level of executive constraints -- I include the
number of peace years since the last conflict incident.[^time] To
account for spillover effects, I include a dummy variable for whether
a neighbouring country is currently experiencing a civil
conflict.[^geo] Finally, one notable lack is a general measure of
political instability, which I omit since it could be highly
correlated with executive constraints and would potentially bias
results, at least for years of crisis where the risk of civil conflict
is highest. I do, however, include a dummy variable for the
independence year since newly sovereign states may be especially weak.

[^geo]: Country neighbours for each year were determined according to
    CShape. Pre-independence conflicts were considered as neighbouring
    to the countries bordering the territory location of the conflict,
    not the government location.

[^time]: To be more specific, civil conflicts can be highly
    destabilizing and severely affect the governing and military
    capacity of a state, increasing the risk for an additional
    conflict onset. Moreover, in times of particularly intense
    conflict, wartime powers may be granted to the executive bypassing
    normal institutional constraints. Thus, ongoing conflict is a
    potential confounder since it can influence both the probability
    of the outcome as well as the executive constraints measure. For
    this purpose, the peace years variable is calculated as the number
    of years since the last conflict-year, including conflict-years
    from extrasystemic conflicts. For countries left censored due to
    the dataset starting date, I set the start of the peace year
    calculation to 1946, the end of World War II. The rationale is
    that World War II was a significant international event that
    reshaped the world order and even had domestic consequences for
    countries not directly involved. For newly independent countries,
    I use the independence date as the start year.

@@

The $\delta$ and $\kappa$ parameters correspond to the slope and
intercept in the linear predictor for the presence of a legislature,
*v2lgbicam*. The effect of their inclusion can be seen in figure
\ref{fig:bicam}. The latent factor, $\theta$, follows the
directionality of the manifest variables, greater values indicate a
more constrained executive and lower values vice versa. The density
plots indicate that on average, for country-years where there is no
legislature, *i.e.* $v2lgbicam$ is zero, the latent factor is lower,
although still conditional on the values for the remaining manifest
variables. Where the legislature is present, $\theta$ is distributed
according to the $Normal(0, 1)$ prior.
